Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-19 paper-new/appendix.tex
work on abstract model constructions

\begin{lemma}\label{lem:step-1-model-to-step-2-model}
Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model}.
Suppose we are given the following data:

\begin{enumerate}
    \item For each value type $A$, a monoid $\pv_A$ and homomorphism 
    \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]

    \item For each computation type $B$, a monoid $\pv_B$ and homomorphism
    \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]

    \item For each value type $A$, a distinguished endomorphism
    $\delta_A \in \ef(FA, FA)$ such that $\delta_A \bisim \id_{FA}$.
\end{enumerate}

Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model}.
\end{lemma}
\begin{proof}
    Write 
    %
    \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
    %
    Define a step-2 model as follows:
    \begin{itemize}
      \item Value objects are tuples of an object $A$ in $\vf$ along with the monoid
      $\pv_A$ and homomorphism $\ptbv_A$:
      $\ob(\vf') = \{ (A, \pv_A, \ptbv_A) \mid A \in \ob(\vf) \}$.

      \item Morphisms are given by morphisms of the underlying objects in $\vf$, i.e.,
       $\vf'((A, \pv_A, \ptbv_A), (A', \pv_{A'}, \ptbv_{A'})) = \vf(A, A')$.

      \item Computation objects are tuples 
      $\ob(\ef') = \{ (B, \pe_B, \ptbe_B) \mid B \in \ob(\ef) \}$.

      \item Computation morphisms are $\ef'((B, \pv_B, \ptbv_B), (B', \pv_{B'}, \ptbv_{B'})) = \ef(B, B')$.

      \item The objects $\vsq'$ and $\esq'$ are the same as those of $\vsq$ and $\esq$.

      \item The morphisms of $\vsq'$ and $\esq'$ are the same as those of $\vsq$ and $\esq$.
    %   \item $\ob(\vsq') = \ob(\vsq)$
    %   \item $\ob(\esq') = \ob(\esq)$
    %   \item $\vsq'(c, c') = \vsq(c, c')$
    %   \item $\esq'(d, d') = \esq(d, d')$

      % Functors \times, +, F, U, arrow
      \item We define $F$ on objects by $F (A, \pv_A, \ptbv_A) = (FA, (1 + \pv_A), h_F)$
      where $1$ is the trivial monoid, $+$ is the coproduct in the category of monoids, and $h_F$ is the homomorphism defined as follows:

      \item We define $U$ on objects by $U (B, \pe_B, \ptbe_B) = (UB, \pe_B, h_U)$
      where $h_U(p_B) = U(\ptbe_B(p_B))$.

      \item We define $(A, \pv_A, \ptbv_A) \arr (B, \pe_B, \ptbe_B) = (A \arr B, \pv_A \times \pe_B, h_\arr)$
      where $\times$ is the product in the category of monoids, and $h_\arr$ is defined by 
      $h_\arr(p_A, p_B) = \ptbv_A(p_A) \arr \ptbe_B(p_B)$.
    \end{itemize}
\end{proof}

@@

\begin{lemma}\label{lem:step-2-model-to-step-3-model}
    Let $\mathcal M$ be a \hyperref[def:step-2-model]{step-2 intensional model}.
    Suppose we are given the following data:

    Then we can construct a \hyperref[def:step-3-model]{step-3 intensional model}.
\end{lemma}
\begin{proof}

\end{proof}

>>>>

\begin{lemma}\label{lem:step-1-model-to-step-2-model}
Let $\mathcal M$ be a \hyperref[def:step-1-model]{step-1 intensional model} with dyn.

Then we can construct a \hyperref[def:step-2-model]{step-2 intensional model} with dyn.
\end{lemma}
\begin{proof}
    % Write 
    % %
    % \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
    % %

    Define a step-2 model $\mathcal M'$ as follows:
    \begin{itemize}
      \item Value objects are tuples consisting of:
      \begin{itemize}
        \item A value object $A$ in $\vf$ 
        \item A monoid of ``pure'' perturbations $P_A$ 
        \item A homomorphism of monoids $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$
        \item A monoid of ``impure'' perturbations $P^K_A$ that contains a distinguished element $\delta^*$
        \item A homomorphism of monoids $\ptbk_A : P^K_A \to \{ \phi \in \ef(FA, FA) \mid \phi \bisim \id_{FA} \}$
        such that $\ptbk_A(\delta^*) = \delta_A^*$
      \end{itemize}

      \item Computation objects are tuples consisting of:
      \begin{itemize}
        \item A computation object $B$ in $\ef$
        \item A monoid of ``pure'' perturbations $P_B$
        \item A homomorphism of monoids $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$
        \item A monoid of ``impure'' perturbations $P^K_B$
        \item A homomorphism of monoids $\ptbk_B : P^K_B \to \{ g \in \vf(UB, UB) \mid g \bisim \id_{UB} \}$.
      \end{itemize}

      \item Morphisms are given by morphisms of the underlying objects in $\vf$ and $\ef$, respectively
      %, i.e.,
      % \[ \vf'((A, P_A, \ptb_A, P^K_A, \ptbk_A), (A', P_{A'}, \ptb_{A'}, P^K_{A'}, \ptbk_{A'})) = \vf(A, A') \]
      %
      % and likewise for computations.

    \end{itemize}

    Before introducing the relations, we make a definition.

    \begin{definition}[push-pull structure]
      Let $c : A \rel A'$ be a value relation of $\mathcal M$. A \emph{value push-pull structure} $\piv_c$ for $c$ consists of:
      \begin{itemize}
        \item A function $\push : P_A \to P_{A'}$ 
              such that for all $\delta^l \in P_A$ we have $\delta^l \ltdyn_c^c \push(\delta^l)$.
        \item A function $\push^K : P^K_A \to P^K_{A'}$ 
              such that for all $\delta^K_l \in P^K_A$ we have $\delta^K_l \ltdyn_{Fc}^{Fc} \push(\delta^K_l)$.
        \item A function $\pull : P_{A'} \to P_A$
              such that for all $\delta_r \in P_{A'}$ we have $\pull(\delta^r) \ltdyn_{c}^c \delta^r$.
        \item A function $\pull^K : P^K_{A'} \to P^K_A$
              such that for all $\delta^K_r \in P^K_{A'}$ we have $\pull(\delta^K_r) \ltdyn_{Fc}^{Fc} \delta^K_r$.
      \end{itemize}

      For $d : B \rel B'$ a computation relation, we define a \emph{computation push-pull structure} $\pie_d$ for $d$
      in an analogous manner.
    \end{definition}

    Now we continue with the description of the construction:
    \begin{itemize}

      \item The objects of $\vsq'$ (i.e., the value relations) are pairs consisting of:
      \begin{itemize}
        \item A value relation $c \in \vsq$
        \item A push-pull structure $\piv_c$ for $c$
      \end{itemize}

      The objects of $\esq'$ are defined analogously.

      \item The morphisms of $\vsq'$ and $\esq'$ are given by the morphisms of $\vsq$ and $\esq$.

      % Functors \times, +, F, U, arrow

      % \item We define $F$ on objects by $F (A, \pv_A, \ptbv_A) = (FA, (1 + \pv_A), h_F)$
      % where $1$ is the trivial monoid, $+$ is the coproduct in the category of monoids, and $h_F$ is the homomorphism defined as follows:

      % \item We define $U$ on objects by $U (B, \pe_B, \ptbe_B) = (UB, \pe_B, h_U)$
      % where $h_U(p_B) = U(\ptbe_B(p_B))$.

      % \item We define $(A, \pv_A, \ptbv_A) \arr (B, \pe_B, \ptbe_B) = (A \arr B, \pv_A \times \pe_B, h_\arr)$
      % where $\times$ is the product in the category of monoids, and $h_\arr$ is defined by 
      % $h_\arr(p_A, p_B) = \ptbv_A(p_A) \arr \ptbe_B(p_B)$.
    \end{itemize}
\end{proof}

@@

\begin{lemma}\label{lem:step-2-model-to-step-3-model}
  Let $\mathcal M$ be a \hyperref[def:step-2-model]{step-2 intensional model}.

  Then we can construct a \hyperref[def:step-3-model]{step-3 intensional model}.
\end{lemma}
\begin{proof}
  Write 
  %
  \[ \mathcal M = (\vf, \vsq, \ef, \esq, \Ff, \Fsq, \Uf, \Usq, \arrf, \arrsq). \] 
  %

  We begin with a definition.

  \begin{definition}[representation structure]
  Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure} $\rho^L_c$ for $c$ consists of
  a value morphism $e_c \in \vf(A, A')$ such that $c$ is quasi-left-representable by $e_c$ (see Definition \ref{def:quasi-left-representable}).

  Likewise, let $d : B \rel B'$. A \emph{right-representation structure} $\rho^R_d$ for $d$ consists of
  a computation morphism $p_d \in \ef(B', B)$ such that $d$ is quasi-right-representable by $p_d$ (see Definition \ref{def:quasi-right-representable}).
  \end{definition}

  (Notice that the direction of the morphism is opposite in the definition of right-representation structure.)

  We define a step-3 model $\mathcal M'$ as follows:
  \begin{itemize}
    \item The objects of $\mathcal M'$ are defined to be the same as the objects of $\mathcal M$.
    \item The value and computation morphisms in $\mathcal M'$ are the same as those of $\mathcal M$.
    \item A value relation is defined to be a tuple $(c, \rho^L_c, \rho^R_{Fc})$ where:
    \begin{itemize}
      \item $c$ is a value relation in $\mathcal M$, and 
      \item $\rho^L_c$ a left-representation structure for $c$, and 
      \item $\rho^R_{Fc}$ a right-representation structure for $Fc$.
    \end{itemize}
    \item Likewise, a computation relation is defined to be a tuple $(d, \rho^R_d, \rho^L_{Ud})$ with
    \begin{itemize}
      \item $d$ a computation relation in $\mathcal M$
      \item $\rho^R_d$ a right-representation structure for $d$
      \item $\rho^L_{Ud}$ a left-representation structure for $Ud$
    \end{itemize}
    \item Morphisms of value relations (i.e., the value squares) are defined by simply
    ignoring the representation structures. That is, a morphism of value relations
    $\alpha \in \vsq'((c, \rho^L_c, \rho^R_{Fc}), (c' \rho^L_{c'}, \rho^R_{Fc'}))$ is simply a morphism of value
    relations in $\vsq(c, c')$. Likewise for computations.
  \end{itemize}
\end{proof}

% Now we define the functors $F$, $U$, $\times$, and $\arr$.

% On objects, the behavior is the same as the respective functors in $\mathcal M$.

% For relations, we define 
% $\Fsq' (c, \rho^L_c, \rho^R_{Fc}) = (\Fsq c, \rho^R_{Fc}, UF(\rho^L_c))$ and
% $\Usq' (d, \rho^R_d, \rho^L_{Ud}) = (\Usq d, \rho^L_{Ud}, FU(\rho^R_d))$.

% We define $(c, \rho^L_c) \arr (d, \rho^R_d) = (c \arr d, \rho^R_{c \arr d})$.

% We now verify that the construction meets the requirements of a step-3 model.
% First, we check that composition of value relations (resp. computation relations)
% is well-defined.

### 2024-01-19 paper-new/categorical-model.tex
work on abstract model constructions

Lastly, we require that for any value object $A$, the hom-set $\ef(FA, FA)$ contains a
distinguished morphism $\delta_A$, such that $\delta_A \bisim_{FA, FA} \id_{FA}$.

@@

\begin{definition}\label{def:step-1-model}
A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
with the additional categories and functors for bisimiarity described above, as  well as
the existence of a distinguised computation morphism $\delta_A \bisim \id_{FA}$ for each $A$.
\end{definition}

@@

We require that $\delta_A \in \pe_{FA}$ for all $A$, where $\delta_A$ is the distinguished
morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition
of a step-1 model.

>>>>

Lastly, we require that for any value object $A$, the hom-set $\ef(FA, FA)$ contains a
distinguished morphism $\delta_A^*$, such that $\delta_A^* \bisim_{FA, FA} \id_{FA}$.

@@

\begin{definition}\label{def:step-1-model}
A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
with the additional categories and functors for bisimiarity described above, as  well as
the existence of a distinguised computation morphism $\delta_A^* \bisim \id_{FA}$ for each $A$.
\end{definition}

@@

We require that $\delta_A^* \in \pe_{FA}$ for all $A$, where $\delta_A^*$ is the distinguished
morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition
of a step-1 model.
